Effects of rotation and magnetic braking on Li abundances for solar-type stars

### 2019-06-24 project/overleaf/paper.tex
Paper v2.3

% Abstract of the paper
\begin{abstract}
\textit{Goals.} The study of the influence of rotation and magnetic braking (MB) on lithium (Li) depletion during pre-main sequence (PMS) and main sequence (MS) for solar-type stars.
\newline\textit{Methods.} The effects of rotational mixing and of the hydrostatic effects of rotation on Li abundances are studied by computing rotating PMS \& MS models. The impact of the MB effect is then analyzed by comparing the Li abundances of those models experiencing different magnetic field intensities ranging between 3.5 and 5.5 G.
\newline\textit{Results.} The surface Li abundance at the end of the PMS and throughout the MS is decreased when rotational effects are included. This results in an increment in the Li depletion rate for rotating models compared to non-rotating ones. This effect is attenuated when the braking effect produced by a magnetic field is added. This physical phenomena, generally neglected in star evolutionary models, impacts also the effective temperature ($T_{eff}$) of the star and therefore its location in the HR diagram. The impact of MB in Li depletion is sensitive to the field intensity: the higher it is, the lower the Li destruction. A relationship between the magnetic fields and the convective zone (CZ) size is observed: stronger magnetic fields produce shallower CZ's. Another important result of this study is that MB effect must be taken into consideration during PMS must if we aim to reproduce Li abundances in young clusters.
\end{abstract}

@@

The denominated standard models, models that include convection only as a mixing process and do not consider other transport options like diffusion or angular momentum loss (AML), have been mainly involved in the elaboration of these predictions \citep{Sestito2005}. Li burns at a temperature $T_{Li} \approx 2.5 x 10^6\; K$, and is therefore directly destroyed in stellar envelopes when the temperature at the base of the convection zone (BCZ) reaches that value. Solar-type stars are characterized by having a CZ that covers much of the stellar radius during the PMS which causes that their lower limit to exceed $T_{Li}$ \citep{Iben1965}. This depletion stops during the approach to the zero-age MS (ZAMS), when the convection zone retreats and the temperature at BCZ is cooler than $T_{Li}$. In the standard stellar models only the mass and the initial chemical composition determine at what distance from the stellar surface $T_{Li}$ temperature is reached, therefore stars within clusters with similar mass are expected to reach the ZAMS with equal surface Li abundances, and remain nearly fixed until the terminal-age MS (TAMS). During this same period the convection mechanisms unleash a mixing process that homogenizes the chemical composition of the convective envelope, from its lower limit to the star surface. However different abundances of Li have been observed for different star populations (\citet{Somers2014} and references therein).\par

@@

The impact of rotation on PMS Li depletion of solar-type stars has already been discussed (\citet{Pinsonneault1997}, \citet{Jeffries2004}, \citet{Somers2014}). These previous studies focused mainly on the hydrostatic effects and didn't consider the influence of a coupling between the star magnetic field and its possible spin-down effect. As previously mentioned, it's thought that AML has a direct influence in the mixing processes and transport of the momentum could be produced by a number of mechanisms that could be important: mass loss, magnetic fields and waves (g-modes). Concerning the two latter, waves and magnetic fields have the property that they generally transmit angular momentum (AM) much more effectively than they induce mixing \citep{Denissenkov2007}. As result, the efficiency of these processes is most easily tested by their impact on AM evolution. Of particular interest for this work is the impact caused by the magnetic field of the star in the AML, more specifically the effect produced on it by MB as a consequence of the solidary rotation that produces on the particles that constitute the solar wind and the torque derived from this coupling. These magnetic fields would originate in the limits between the radiative and convective zones of the star, in the so-called tachoclines, due to the rotational difference between both zones \citep{Guerrero2016}. A proper consideration of the interaction between rotation and magnetic fields is therefore essential. We have adapted the one-dimensional stellar evolution code used in this work to model MB and include its effect on the equations for the evolution of the AM distribution and stellar structure. This produces a more complete, though still reasonably simple, model for AM evolution. In this study we are now focused in investigating not only the effects of rotation on the Li content during PMS and MS but also on the indirect role played by the MB on the Li destruction as a result of its influence on the rotational history of solar-type stars.\par

@@

We define here also some conventions and assumptions adopted in the paper. For the models presented in this paper we adopt solar-scaled abundances and assumes the \citet{Asplund2009} protosolar birth cloud metallicity, not the current photospheric metallicity, as the reference value. In other words, [Z/H] is computed with respect to $Z = Z_{\odot, protosolar} = 0.0142$, not $Z = Z_{\odot, photosphere} = 0.0134$. Lastly, in accordance with the XXIXth IAU resolutions B3 \citep{Mamajek2015} we adopt the following nominal values to express stellar properties values in SI units $R_{\odot} = 6.957x10^{10} cm$ and $M_{\odot} = 1.988x10^{33} g$. For a detailed description of the physics adopted in this paper, refer to \citet{Choi2016}. This work has been used as a starting point for our work regarding the parameterization of the MESA project calibrated to reproduce the measured values on the solar surface, in particular the section 4.1 of the aforementioned paper.\par

The paper is organized as follows. Section 2 gives an overview of rotation and magnetic fields physical processes and how do they impact on the star evolution. We also summarized the most relevant parameters configured in MESA for the model simulations. In Section 3, the model outputs are presented and discussed in detail. Finally, Section 4 finishes this paper with a discussion of open points and future works.\par

@@

In this section we review the core physic aspects simulated in MESA and the most relevant parameters adopted in the models simulations. In its simplest version, by activating rotation in the model, the evolutionary computational code MESA takes into account the following effects:
%\vspace{-1pc}
\begin{enumerate}
    \item AM transport from the radiative interior to the convective envelope in response to the rotational deceleration of the stellar surface layers\label{itm:1}.
    \item AM redistribution associated with changes in internal structure during the process of contraction to the MS\label{itm:2}.
    \item AML as a result of the torque applied to the convection zone by a magnetically coupled wind\label{itm:3}.
\end{enumerate}

@@

With regards to MESA, the transport of AM and chemical elements due to rotationally induced instabilities is implemented in a diffusion approximation \citet{Endal1978}. MESA calculates diffusion coefficients for five different rotationally induced mixing processes: dynamical shear instability, Solberg-H{\o}iland instability, secular shear instability, Eddington-Sweet circulation, and the Goldreich-Schubert-Fricke instability \citep{Paxton2013}.

\subsection{Modelling magnetic braking}
The origin of magnetic fields is still unknown. The organized magnetic fields detected in some O stars \citep{Wade2010} might be fossil fields (see \citet{Dudorov2014} for details about the Theory of fossil magnetic field), or fields produced through a dynamo mechanism \citep{Cantiello2009}. Besides that, it has long been a requirement for T Tauri stars to host kG magnetic fields \citep{Hussain2014} in order to explain several key observational characteristics in young, accreting PMS systems \citep{Johns-Krull2007}. One of this features is the long rotating periods  which is much longer than predicted from AM conservation.  A key question is to understand how these young stars can accrete large amounts of material from their circumstellar disk with high specific angular momentum and at the same time they're able to keep slow rotation periods, between 7-10 days \citep{Hussain2014}. One plausible explanation for this behaviour considers the AML produced by the interaction between the magnetic field and the particles that form its circumstellar disk \citep{Zanni2012}. Once accretion stops or becomes less efficient then stars are free to spin up as they contract towards the MS.\par

@@

\begin{figure}
	\includegraphics[width=\columnwidth]{figures/hussain_gaitee_fig1.png}
    \caption{H-R diagram showing Behrend \& Maeder (2001) pre-main sequence evolutionary tracks for stellar masses up to $4.0 \; M_{\sun}$. The dashed blue line marks the position of the birthline. All stars with $1.0 \; M_{\sun}$ will undergo a stage along their PMS evolution in which they have a fully convective interior near the birthline, to developing a radiative core when it's approaching the ZAMS (black dot-dashed line) \citet{Hussain2014}}
    \label{fig:hr_hussain}
\end{figure}

@@

It is also important to emphasize that our models fail to match the observed solar Li abundance and $\Omega$. We cannot, therefore, ascertain to have correctly modeled the rotational history of the Sun. In addition, we cannot (yet) properly explain the ups and downs observed in the $\Omega$ at the end of the MS. Therefore, in view of these shortcomings in our models, we must analyse the results obtained with caution and not draw any premature conclusions. Having said that, we are convinced that our research is on the right track and we do believe that the following conclusions, based on the results described in previous sections, have been sufficiently supported in the course of this work:
\begin{enumerate}
    \item The $T_{eff}$ from standard evolutionary tracks represent upper limits since these models do not take into consideration the magnetic braking effect.
    \item The combination of rotation during the PMS and MB effect during MS produces different, potentially more promising behavior than those produced by standard models.
    \item Inclusion of the magnetic field leads to cooler models and lower Li depletions in the MS.
    \item MB during PMS seems to be also required for explaining Li abundances in young clusters.
\end{enumerate}

>>>>

% Abstract of the paper
\begin{abstract}
\textit{Goals.} The study of the influence of rotation and magnetic braking (MB) on lithium (Li) depletion during pre-main sequence (PMS) and main sequence (MS) for solar-type stars.
\newline\textit{Methods.} The effects of rotational mixing and of the hydrostatic effects of rotation on Li abundances are studied by simulating several grids of PMS \& MS rotating and non-rotating models. The previous effects are combined with the additional impact of the MB one (with magnetic field intensities ranging between 3.5 and 5.5 G) and the data obtained from the two set of simulations are confronted by comparing a number of parameters: Li abundances, angular velocity, effective temperature and size of convective zone.
\newline\textit{Results.} The surface Li abundance at the end of the PMS and throughout the MS is decreased when rotational effects are included, i.e. the Li depletion rate for rotating models is higher than for non-rotating ones. However, this effect is attenuated when the braking effect produced by a magnetic field is present. This physical phenomena, generally neglected in star evolutionary models, impacts also the effective temperature ($T_{eff}$) of the star and therefore its location in the HR diagram. The impact of MB in Li depletion is sensitive to the field intensity: the higher it is, the lower the Li destruction. A direct link between the magnetic fields and the convective zone (CZ) size is observed: stronger magnetic fields produce shallower CZ's. Another important result of this study is that MB effect must be taken into consideration during PMS must if we aim to reproduce Li abundances in young clusters.
\end{abstract}

@@

The denominated standard models, models that include convection only as a mixing process and do not consider other transport options like diffusion or angular momentum loss (AML), have been mainly involved in the elaboration of these predictions \citep{Sestito2005}. Li is destroyed at a temperature $T_{Li} \approx 2.5 x 10^6\; K$, and is therefore directly destroyed in stellar envelopes when the temperature at the base of the convection zone (BCZ) reaches that value. Solar-type stars are characterized by having a CZ that covers much of the stellar radius during the PMS which causes that their lower limit to exceed $T_{Li}$ \citep{Iben1965}. This depletion stops during the approach to the zero-age MS (ZAMS), when the convection zone retreats and the temperature at BCZ is cooler than $T_{Li}$. In the standard stellar models only the mass and the initial chemical composition determine at what distance from the stellar surface $T_{Li}$ temperature is reached, therefore stars within clusters with similar mass are expected to reach the ZAMS with equal surface Li abundances, in addition to showing a very similar Li evolution until their approach to the terminal-age MS (TAMS). During this same period the convection mechanisms unleash a mixing process that homogenizes the chemical composition of the convective envelope, from its lower limit to the star surface. However different abundances of Li have been observed for different star populations (\citet{Somers2014} and references therein).\par

@@

The impact of rotation on PMS and on Li depletion for solar-type stars has already been discussed (\citet{Pinsonneault1997}, \citet{Jeffries2004}, \citet{Somers2014}). These previous studies focused mainly on the hydrostatic effects and didn't consider the influence of a coupling between the star magnetic field and its possible spin-down effect. As previously mentioned, it's thought that AML has a direct influence in the mixing processes and transport of the momentum could be produced by a number of mechanisms that could be important: mass loss, magnetic fields and waves (g-modes). Concerning the two latter, waves and magnetic fields have the property that they generally transmit angular momentum (AM) much more effectively than they induce mixing \citep{Denissenkov2007}. As result, the efficiency of these processes is most easily tested by their impact on AM evolution. Of particular interest for this work is the impact caused by the magnetic field of the star in the AML, more specifically the effect produced on it by MB as a consequence of the solidary rotation that produces on the particles that constitute the solar wind and the torque derived from this coupling. These magnetic fields would originate in the limits between the radiative and convective zones of the star, in the so-called tachoclines, due to the rotational difference between both zones \citep{Guerrero2016}. A proper consideration of the interaction between rotation and magnetic fields is therefore essential. We have adapted the one-dimensional stellar evolution code used in this work to model MB and include its effect on the equations for the evolution of the AM distribution and stellar structure. This produces a more complete, though still reasonably simple, model for AM evolution. In this study we are now focused in investigating not only the effects of rotation on the Li content during PMS and MS but also on the indirect role played by the MB on the Li destruction as a result of its influence on the rotational history of solar-type stars.\par

@@

We define here also some conventions and assumptions adopted in the paper. For the models presented in this paper we adopt solar-scaled abundances and assumes the \citet{Asplund2009} protosolar birth cloud metallicity ($Z = Z_{\odot, proto} = 0.0142$, not the current photospheric metallicity $Z = Z_{\odot, photosphere} = 0.0134$, as the reference value. We also adopt the following nominal values to express stellar properties values in SI units $R_{\odot} = 6.957x10^{10} cm$ and $M_{\odot} = 1.988x10^{33} g$ which are consistent with IAU resolutions \citep{Mamajek2015} . For a detailed description of the physics adopted in this paper, refer to \citet{Choi2016}. This work has been used as a starting point for our work regarding the parameterization of the MESA project calibrated to reproduce the measured values on the solar surface, in particular the section 4.1 of the aforementioned paper.\par

This paper is structured as follows: Section 2 gives an overview of rotation and magnetic fields physical processes and how do they impact on the star evolution. We also summarized the most relevant parameters configured in MESA for the model simulations. In Section 3, the model outputs are presented and discussed in detail. Finally, Section 4 finishes this paper with a discussion of open points and future works.\par

@@

In this section we review the core physic aspects simulated in MESA and the most relevant parameters adopted in the models simulations. In its simplest version, by activating rotation in the model, the evolutionary computational code MESA takes into account the following effects:
%\vspace{-1pc}
\begin{enumerate}
    \item AM transport from the radiative interior to the convective envelope in response to the rotational deceleration of the star surface layers\label{itm:1}.
    \item AM redistribution associated with changes in internal structure during the process of contraction to the MS\label{itm:2}.
    \item AML as a result of the torque applied to the convection zone by a magnetically coupled wind\label{itm:3}.
\end{enumerate}

@@

In reference to MESA and as stated in \citet{Paxton2013}: "the transport of AM and chemical elements due to rotationally induced instabilities is implemented in a diffusion approximation \citet{Endal1978}. MESA calculates diffusion coefficients for five different rotationally induced mixing processes: dynamical shear instability, Solberg-H{\o}iland instability, secular shear instability, Eddington-Sweet circulation, and the Goldreich-Schubert-Fricke instability".

\subsection{Modelling magnetic braking}
The origin of magnetic fields is still unknown. The organized magnetic fields detected in some O stars \citep{Wade2010} might be fossil fields (see \citet{Dudorov2014} for details about the Theory of fossil magnetic field), or fields produced through a dynamo mechanism \citep{Cantiello2009}. Besides that, the presence of magnetic fields with intensities of the order of kG \citep{Hussain2014} has been a sine qua non condition for dealing with certain data observed in T Tauri type stars in young, accreting PMS systems \citep{Johns-Krull2007}. One of this features is the long rotating periods  which is much longer than predicted from AM conservation. A key question is to understand how these young stars can accrete large amounts of material from their circumstellar disk with high specific angular momentum and at the same time they're able to keep slow rotation periods, between 7-10 days \citep{Hussain2014}. One plausible explanation for this behaviour considers the AML produced by the interaction between the magnetic field and the particles that form its circumstellar disk \citep{Zanni2012}. Once the accretion process stops or becomes less efficient then the star is able to spin up as it contracts towards the ZAMS.\par

@@

\begin{figure}
	\includegraphics[width=\columnwidth]{figures/hussain_gaitee_fig1.png}
    \caption{H-R diagram showing PMS evolutionary tracks for stars with masses up to $4.0 \; M_{\sun}$. The dashed blue line marks the position of the birthline. The stars with $1.0 \; M_{\sun}$ which have a fully convective interior near the Hayashi track, they will to develop a radiative core when they're approaching the ZAMS (black dot-dashed line). Image copied from \citet{Hussain2014}}
    \label{fig:hr_hussain}
\end{figure}

@@

It is also important to emphasize that our models fail to match the observed solar Li abundance and $\Omega$. We cannot, therefore, ascertain to have correctly modeled the rotational history of the Sun. In addition, we cannot (yet) properly explain the ups and downs observed in the $\Omega$ at the end of the MS. Therefore, in view of these shortcomings in our models, we must analyse the results obtained with caution and not draw any premature conclusions. Having said that, we are convinced that our research is on the right track and we do believe that the following conclusions, based on the results described in previous sections, have been sufficiently supported in the course of this work:
\begin{enumerate}
    \item Inclusion of the magnetic field leads to cooler models and lower Li depletions in the MS.
    \item A combination of rotation during the PMS and MB effect during MS produces different, potentially more promising behavior than those produced by standard models.
    \item $T_{eff}$ from standard evolutionary tracks represent upper limits since these models do not take into consideration the magnetic braking effect.
    \item The convective zone extension decreases when the intensity of the magnetic field increases.
    \item MB during PMS seems to be also required for explaining Li abundances in young clusters.
\end{enumerate}
